---
title: 'Appendix III'
subtitle: 'Ecological memory patterns of virtual taxa: analyses and results'
author: "Blas M. Benito"
geometry: "left=1cm,right=1cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    fig_caption: yes
    fig_width: 9
    highlight: tango
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 2
    df_print: kable
  html_notebook:
    fig_caption: yes
    highlight: haddock
    toc: yes
    toc_depth: 3
code_folding: show
citation_package: natbib
---

<!-- knit_hooks$set(document  =  function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)}) -->

<!-- main info on pdf and notebooks -->
<!-- http://rmarkdown.rstudio.com/pdf_document_format.html#latex_engine -->

<!-- check xelatex fonts -->
<!-- fc-list : family -->
<!-- fc-list :fontformat = TrueType -->

<!-- Equations cheatsheet -->
<!-- $a_{b}$ -->
<!-- $a^{b}$ -->
<!-- $\alpha, \beta, \ldots$ -->
<!-- $\sum_{n = 1}^N$ -->
<!-- $\times$ -->
<!-- $\prod_{n = 1}^N$ -->
<!-- $<, \leq, \geq$ -->
<!-- $\sim$ -->
<!-- $\widehat{\alpha}$ -->
<!-- $\bar{x}$ -->
<!-- $\frac{a}{b}$ -->
<!-- $\left(\frac{a}{b}\right)$ -->


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#######################################################################
options(scipen=999)

#checking if required packages are installed, and installing them if not
#checking if required packages are installed, and installing them if not
list.of.packages <- c("ggplot2", "cowplot", "knitr", "viridis", "tidyr", "formatR", "grid", "zoo", "ranger", "rpart", "rpart.plot", "partykit", "HH", "pdp", "kableExtra", "magrittr", "stringr", "dplyr", "grid", "gridExtra", "virtualPollen")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, dep = TRUE)

#from Github
if(!("virtualPollen" %in% installed.packages())){
  library(devtools)
  install_github("blasbenito/memoria")
}

library(virtualPollen)
library(memoria)
library(ggplot2) #plotting library
library(cowplot) #plotting library
library(knitr)   #report generation in R
library(viridis) #pretty plotting colors
library(grid)    #plotting 
library(tidyr)
library(formatR)
library(zoo)     #time series analysis
library(HH)      #variance inflation factor (multicollinearity analysis)
library(kableExtra) #to fit tables to pdf page size
library(magrittr) #kableExtra requires pipes
library(pdp)     #partial dependence plots
library(ranger)  #fast Random Forest implementation
library(rpart)   #recursive partitions trees
library(rpart.plot) #fancy plotting of rpart models
library(stringr) #to parse variable names
library(gridExtra)
library(grid)

options(scipen = 999)

# setting code font size in output pdf, from https://stackoverflow.com/a/46526740
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

#trying to line-wrap code in pdf output
#from https://github.com/yihui/knitr-examples/blob/master/077-wrap-output.Rmd
knitr::opts_chunk$set(echo = TRUE, fig.pos = "h")
  opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = FALSE)

#loading output of the previous appendix
load("Appendix_I_output.RData")
load("Appendix_III_output.RData")

rm(list.of.packages, new.packages)
```

&nbsp;

**Summary**

This appendix presents the code required to run the analyses and to generate the figures presented in the paper. The aim of this appendix is to ensure the reproducibility of our work. To reach a full understanding of what the code in this appendix is doing requires to at least read the paper, and go through the contents of **Appendix II**. Please bear in mind that the code shown in this appendix can take hours to execute.

**IMPORTANT:** An Rmarkdown version of this document can be found at: https://github.com/BlasBenito/EcologicalMemory.

\pagebreak

# Quantifying ecological memory patterns for the simulated taxa and dataset types

We evaluated ecological memory on 16 virtual taxa and 5 dataset types ("Annual", "1cm", "2cm", "6cm", and "10cm") by using separatedly as exogenous component the values of the *driver* and the *suitability* values returned by the niche functions of the taxa. To compute ecological memory patterns from the simulated data, these steps are required:


+  Run Random Forest models, as explained in **Appendix II** on each combination of taxa and dataset type by using the *runExperiment* function.
+  Plot the outcome of *runExperiment* to examinate results visually.
+  Organize the results into a single table with *experimentToTable*.

After these steps, results are ready for further analyses. 

## Fitting Random Forest models

General parameters for the simulation.

```{r, cache=FALSE, size="small", echo=FALSE}
#dataset types and taxa to be analyzed
selected.dataset.types = 1:5
selected.taxa = 1:16

#number of Random Forest models to fit
repetitions = 300

#names of these columns
sampling.names = c("Annual", 
                   "1cm", 
                   "2cm", 
                   "6cm", 
                   "10cm")

#response variable
response.column = "Pollen"

#time column
time.column = "Time"

#lags
lags = seq(20, 240, by=20)

#species traits to include in the output
traits <- c("maximum.age",
            "fecundity", 
            "niche.A.mean",
            "niche.A.sd")
```


Using **suitability** as exogenous component.

```{r, size="small", message=FALSE, warning=FALSE, cache=FALSE, cache=FALSE, eval=FALSE}
E1.suitability <- runExperiment(
  simulations.file = simulation,
  selected.rows = selected.taxa,
  selected.columns = selected.dataset.types,
  parameters.file = parameters,
  parameters.names = traits,
  sampling.names = sampling.names,
  driver.column = "Suitability",
  response.column = response.column,
  subset.response = "none",
  time.column = time.column,
  time.zoom = NULL,
  lags = lags,
  repetitions = repetitions)

```

\pagebreak

Using **driver** as exogenous component.

```{r, size="small", message=FALSE, warning=FALSE, cache=FALSE, eval=FALSE}
E1.driver <- runExperiment(
  simulations.file = simulation,
  selected.rows = selected.taxa,
  selected.columns = selected.dataset.types,
  parameters.file = parameters,
  parameters.names = traits,
  sampling.names = sampling.names,
  driver.column = "Driver.A", #only difference
  response.column = response.column,
  subset.response = "none",
  time.column = time.column,
  time.zoom = NULL,
  lags = lags,
  repetitions = repetitions)
```


## Plotting ecological memory patterns

The experiments ran in the previous section can be easily plotted with the *plotExperiment()* function as follows:

```{r, size="small", echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
plotExperiment(
  experiment.output=E1.suitability, 
  experiment.title="plot title", 
  legend.position="bottom", 
  R2=TRUE, 
  sampling.names=sampling.names, 
  strip.text.size=7, 
  axis.x.text.size=9, 
  axis.y.text.size=12, 
  axis.x.title.size=14, 
  axis.y.title.size=14,
  title.size=18,
  filename = "filename",
  caption = "figure caption")
```

The result of this function applied on the experiments is shown below.

\newpage

```{r, size="small", echo=FALSE, fig.width=15, fig.height=18.5, message=FALSE, warning=FALSE, cache=FALSE, fig.cap="Ecological memory patterns of all virtual species (rows) and dataset types (columns) when using suitability (blue) as exogenous component (blue curves)."}
  plotExperiment(
  experiment.output=E1.suitability, 
  experiment.title="Ecological memory patterns when using 
  suitability as exogenous component.", 
  legend.position="bottom", 
  R2=TRUE, 
  sampling.names=sampling.names, 
  strip.text.size=7, 
  axis.x.text.size=9, 
  axis.y.text.size=12, 
  axis.x.title.size=14, 
  axis.y.title.size=14,
  title.size=18,
  filename = "Experiment_suitability",
  caption = "Rows represent virtual taxa; Column 1: Annual dataset interpolated at 20 years resolution;  Columns 2 to 5: sedimenta ccumulation rate is applied and are sampled every 1, 2, 6, and 10 cm. respectively")
```

```{r, size="small", echo=FALSE , fig.width=15, fig.height=18.5, message=FALSE, warning=FALSE, cache=FALSE, fig.cap="Ecological memory patterns of all virtual species (rows) and dataset types (columns) when using driver (blue) as exogenous component (blue curves)."}
plotExperiment(
  experiment.output=E1.driver, 
  experiment.title="Ecological memory patterns when using 
  driver as exogenous component.", 
  legend.position="bottom", 
  R2=TRUE, 
  sampling.names=sampling.names,
  strip.text.size=7, 
  axis.x.text.size=9, 
  axis.y.text.size=12, 
  axis.x.title.size=14, 
  axis.y.title.size=14,
  title.size=18,
  filename = "Experiment_driver",
  caption = "Rows represent virtual taxa; Column 1: Annual dataset interpolated at 20 years resolution;  Columns 2 to 5: sedimenta ccumulation rate is applied and are sampled every 1, 2, 6, and 10 cm. respectively")
```

## Organizing results into dataframes to facilitate further analyses

When applying *experimentToTable* to **E1.suitability** or **E1.driver** the resulting columns are (see **Table 1**):


+ **median**: median importance of the given **Variable** at a given **Lag**.
+ **sd**: standard deviation of the importance.
+ **min**: 0.05 percentile of the variable importance across repetitions.
+ **max**: 0.95 percentile of the variable importance.
+ **Variable**: name of the given variable (response, suitability, driver, random).
+ **Lag**: lag ID.
+ **R2mean**: mean pseudo R-squared of Random Forest models across repetitions.
+ **R2sd**: standard deviation of pseudo R-squared.
+ **VIFmean**: average variance inflation factor (VIF) of the predictors for the given dataset.
+ **VIFsd**: standard deviation of VIF values.
+ **label**: name of the virtual taxa as shown in the parameters dataframe.
+ **maximum.age**: maximum life-span of the virtual taxa.
+ **reproductive.age**: reproductive age.
+ **fecundity**: maximum fecundity under ideal conditions.
+ **growth.rate**: growth rate.
+ **maximum.biomass**: maximum biomass of the individuals (100).
+ **carrying.capacity**: carrying capacity of the landscape (10000).
+ **niche.A.mean**: mean of the niche function. Also, *niche optimum* or *niche position*.
+ **niche.A.sd**: standard deviation of the niche function. Also, *niche breadth*.
+ **sampling**: one of "Annual", "1cm", "2cm", "6cm", "10cm"

The code below generates the tables for the **suitability** and **driver** data. A sample output of *experimentToTable* is shown below.

```{r, size="small", message=FALSE, warning=FALSE, cache=FALSE, eval=FALSE}
E1.suitability.df <- experimentToTable(
  experiment.output=E1.suitability, 
  parameters.file=parameters, 
  sampling.names=sampling.names, R2=TRUE)

E1.driver.df <- experimentToTable(
  experiment.output=E1.driver, 
  parameters.file=parameters, 
  sampling.names=sampling.names, R2=TRUE)
```

\newpage
\blandscape
\thispagestyle{empty}

```{r table1, echo = FALSE}
#removing some useless columns
E1.suitability.df$niche.B.mean <- NULL
E1.suitability.df$niche.B.sd <- NULL
E1.suitability.df$autocorrelation.length.A <- NULL
E1.suitability.df$autocorrelation.length.B <- NULL
E1.suitability.df$driver.A.weight <- NULL
E1.suitability.df$driver.B.weight <- NULL
E1.suitability.df$pollen.control <- NULL
E1.suitability.df$name <- NULL

#rounding some columns
E1.suitability.df$median <- round(E1.suitability.df$median, 2)
E1.suitability.df$sd <- round(E1.suitability.df$sd, 2)
E1.suitability.df$min <- round(E1.suitability.df$min, 2)
E1.suitability.df$max <- round(E1.suitability.df$max, 2)
E1.suitability.df$R2mean <- round(E1.suitability.df$R2mean, 2)
E1.suitability.df$R2sd <- round(E1.suitability.df$R2sd, 5)
E1.suitability.df$VIFmean <- round(E1.suitability.df$VIFmean, 2)
E1.suitability.df$VIFsd <- round(E1.suitability.df$VIFsd, 5)

rownames(E1.suitability.df) <- NULL

#rendering table
kable(E1.suitability.df[1:35, ], caption="First rows of the table E1.suitability.df", booktabs = T) %>% 
  kable_styling(latex_options = c("scale_down", "hold_position", "striped")) %>%  
  column_spec(1:ncol(E1.suitability.df), color="#38598C") %>%
  row_spec(0, col="#585858")

#saving output
# save(E1.driver, E1.driver.df, E1.suitability, E1.suitability.df, file = "Appendix_III_output.RData")

```

\vfill
\raisebox{0.1cm}{\makebox[\linewidth]{\thepage}}
\elandscape
\newpage

# Paper figures and tables

\setcounter{figure}{0} 
\setcounter{table}{0} 

```{r, echo=FALSE, eval=TRUE}

#GENERAL PARAMETERS FOR FIGURES

#PUBLICATION SIZE
#DIMENSIONS
figure.width=2.75
annotation.size=2
legend.title.size=6
legend.text.size=6
axis.text.size=6
axis.title.size=6

#COLORS
plot.cols <- viridis(12, option="D")
col.blue<-plot.cols[4]
col.green<-plot.cols[9]
col.gray.dark <- "gray20"
col.gray.medium <-"gray50"
col.gray.light <-"gray80"

#listing objects in the environment
to.preserve <- ls()
to.preserve <- c(to.preserve, "to.preserve") #most beautiful line of code ever
# rm(list= ls()[!(ls() %in% to.preserve)]) #to remove all objects except these
```

## Figure 1


```{r, echo=FALSE, fig.height=2, fig.width=2.75, fig.cap="Components of ecological memory. Antecedent values of the driver and the response for a given lag length are located at T – Lag. Arrows represent the relative contribution of endogenous memory, exogenous memory, and the concurrent effect on the value of interest of the response at time T, represented by the green dot. This conceptual structure can be scaled up to any number of drivers and lags."}

#FIGURE 1: MEMORY PATTERNS
set.seed(10)
time <- 1:100
Driver <- seq(1, 10, length.out = 100) + cumsum(sample(c(-0.1, 0, 0.1), max(time), replace=TRUE))

#SUBSET PLOT DATA
set.seed(10)
plot.data<-data.frame(Lag=time, Driver=Driver)
set.seed(10)
plot.data$Response<-jitter(rescaleVector(plot.data$Driver, new.min=30, new.max=90), amount=3) + 8 
set.seed(10)
plot.data$Driver<-jitter(rescaleVector(plot.data$Driver, new.min=10, new.max=40), amount=3)
# plot.data$Driver<-jitter(rescaleVector(plot.data$Driver, new.min=5, new.max=60), amount=0.6)
plot.data$Lag<-round(rescaleVector(plot.data$Lag, new.min=0, new.max=100), 0)

#long format for ggplot2
plot.data<-gather(plot.data, Variable, value, 2:3)

T0<-20
T1<-80

#curve configuration
myCurve1<-curveGrob(0, 0, 1, 1, default.units = "npc",
                    curvature = -0.33, angle = 60, ncp = 10, shape = 1,
                    square = FALSE, squareShape = 1,
                    inflect = FALSE, arrow = arrow(length = unit(0.1, "inches")), open = TRUE,
                    debug = FALSE,
                    name = NULL, gp = gpar(lwd=1, col=col.blue), vp = NULL)

myCurve2<-curveGrob(0, 0, 1, 1, default.units = "npc",
                    curvature = 0.10, angle = 100, ncp = 10, shape = 1,
                    square = FALSE, squareShape = 1,
                    inflect = FALSE, arrow = arrow(length = unit(0.1, "inches")), open = TRUE,
                    debug = FALSE,
                    name = NULL, gp = gpar(lwd=1, col=col.green), vp = NULL)

myCurve3<-curveGrob(0, 0, 1, 1, default.units = "npc",
                    curvature = 0.4, angle = 60, ncp = 10, shape = 1,
                    square = FALSE, squareShape = 1,
                    inflect = FALSE, arrow = arrow(length = unit(0.1, "inches")), open = TRUE,
                    debug = FALSE,
                    name = NULL, gp = gpar(lwd=1, col="gray40"), vp = NULL)

myCurve3b<-curveGrob(0, 0, 1, 1, default.units = "npc",
                     curvature = 0.4, angle = 60, ncp = 10, shape = 1,
                     square = FALSE, squareShape = 1,
                     inflect = FALSE, arrow = arrow(length = unit(0.1, "inches")), open = TRUE,
                     debug = FALSE,
                     name = NULL, gp = gpar(lwd=1, lty=3, col="gray40"), vp = NULL)

myCurve3c<-curveGrob(0, 0, 1, 1, default.units = "npc",
                     curvature = 0.2, angle = 60, ncp = 10, shape = 1,
                     square = FALSE, squareShape = 1,
                     inflect = FALSE, arrow = arrow(length = unit(0.1, "inches")), open = TRUE,
                     debug = FALSE,
                     name = NULL, gp = gpar(lwd=1, lty=3, col=col.green), vp = NULL)

myCurve4<-curveGrob(0, 0, 1, 1, default.units = "npc",
                    curvature = 0, angle = 0, ncp = 10, shape = 1,
                    square = FALSE, squareShape = 1,
                    inflect = FALSE, arrow = NULL, open = TRUE,
                    debug = FALSE,
                    name = NULL, gp = gpar(lwd=1), vp = NULL)

#plot
p<-ggplot(data=plot.data, aes(x=Lag, y=value, colour=Variable)) + 
  geom_line(size=1, alpha=0.8) + 
  ylim(0,100) +
  geom_vline(xintercept=T0) + 
  geom_vline(xintercept=T1) + 
  scale_color_manual(values=c(col.green, col.blue)) +
  scale_x_continuous(breaks=c(T0, T1), labels=c("T - Lag", "T")) +
  annotate("text", x=T0+16, y=94, label="Endogenous", color=col.blue, size=annotation.size) + 
  annotate("text", x=T0+16, y=87, label="memory", color=col.blue, size=annotation.size) +
  annotate("text", x=T0+22, y=52, label="Exogenous", color=col.green, size=annotation.size) + 
  annotate("text", x=T0+22, y=45, label="memory", color=col.green, size=annotation.size) +
  annotate("text", x=T1+16, y=69, label="Concurrent", color=col.gray.medium, size=annotation.size) + 
  annotate("text", x=T1+16, y=62, label="effect", color=col.gray.medium, size=annotation.size) +
  annotate("text", x=T0+30, y=7, label="Lag", color=col.gray.dark, size=annotation.size) +
  annotation_custom(grob=myCurve1,T0,T1-1,49,85+3) +
  annotation_custom(grob=myCurve2,T0,T1-1,17,85-3) + 
  annotation_custom(grob=myCurve3,T1,T1+1,33,85-3) +
  # annotation_custom(grob=myCurve3b,T0,T0+0.5,17,49-1) +
  # annotation_custom(grob=myCurve3c,T0,T1,17,33) +
  annotation_custom(grob=myCurve4,T0+1,T1-1,0,0) +
  geom_point(aes(x=T1, y=85), shape=21, color="gray20", fill=col.blue, size=2, stroke=0.5) +
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(),
        axis.line=element_blank(),
        legend.margin=margin(-1, unit = "cm"),
        legend.text=element_text(size=legend.text.size),
        legend.title=element_text(size=legend.title.size),
        axis.text.x=element_text(size=axis.text.size),
        plot.margin = unit(c(0.2, 0.2, 0.8, 0.2), units = "cm"),
        legend.position = c(0.2, -0.4),
        legend.direction = "horizontal") +
  labs(color="Variables:")
    

print(p)

# invisible(print(grid.arrange(arrangeGrob(p))))
ggsave(filename = "Figure_1.pdf", width=figure.width, height=2)
# ggsave(filename = "/home/blas/Dropbox/RESEARCH/PROJECTS/MAIN_PROJECTS/IGNEX/GITHUB/Ecological-Memory/paper/figures/Figure_1.pdf", width=figure.width, height=2)

#removing objects we don't need any longer
rm(p, myCurve1, myCurve2, myCurve3, myCurve3b, myCurve3c, myCurve4, plot.data, T0, T1)

#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])

```


## Figure (removed from paper)

This figure was generated with the package *yEd* (URL: www.yworks.com/products/yed).

```{r, fig.width=2.75, include=TRUE, fig.align="center", fig.cap="Model structure. Green and white boxes represent respectively the environment and elements relative to the virtual taxa. Light-gray boxes are the experimental variables, changing across model executions. Population dynamics (dashed ellipse) is the emergent process from which pollen production (blue box), the response variable, results.", echo=FALSE}
knitr::include_graphics("Figure_2.pdf")
```

## Figure (removed from paper)

```{r, echo=FALSE, fig.height=3.75, fig.width=2.75, cache=FALSE, fig.cap="Panel a represents the different ecological niches used in the simulations. Light green colour represents the driver and marine blue represents the niche of the virtual taxa. Numbers in grey strips represent the mean (m) and standard deviation (sd) of the Gaussian functions used to define the niche. Panel b shows the dynamics of the driver  during 2000 years, and the suitability values returned by the different niche functions in panel a. Panel c shows the two combinations of life-span used in the simulations, with their respective growth rates, and sexual maturity indicated by a blue vertical line."}

i=1
parameters.plot <- parametersDataframe(rows=1)
parameters.plot[i,] <- c("No 50, Nb 10", 10, 4, 10, 1.5, 0, 100, 10000, 1, 0, 50, 10, 50, 5, 600, 600)
i=i+1
parameters.plot[i,] <- c("No 50, Nb 20", 10, 4, 10, 1.5, 0, 100, 10000, 1, 0, 50, 20, 50, 5, 600, 600)
i=i+1
parameters.plot[i,] <- c("No 75, Nb 10", 1000, 100, 10, 0.01, 0, 100, 10000, 1, 0, 75, 10, 50, 5, 600, 600)
i=i+1
parameters.plot[i,] <- c("No 75, Nb 20", 1000, 100, 10, 0.01, 0, 100, 10000, 1, 0, 75, 20, 50, 5, 600, 600)
i=i+1
parameters.plot[i,] <- c("Growth-rate 1.5", 10, 4, 1, 1.5, 0, 100, 10000, 1, 0, 75, 20, 50, 5, 600, 600)
i=i+1
parameters.plot[i,] <- c("Growth-rate 0.01", 1000, 100, 10, 0.01, 0, 100, 10000, 1, 0, 50, 10, 50, 5, 600, 600)
parameters.plot[, 2:ncol(parameters.plot)] = sapply(parameters.plot[, 2:ncol(parameters.plot)], as.numeric)

#dataframe to store data
plot.df = data.frame(Species=character(), Driver.density.x=numeric(), Driver.density.y=numeric(), Value=numeric(), Suitability=numeric(), Age=numeric(), Biomass=numeric(), Reproductive.age=numeric(), Fecundity=numeric())


#ITERATING THROUGH SPECIES
for(i in 1:nrow(parameters.plot)){
  
  #GETTING DRIVER DATA
  driver.A = na.omit(simulation[[1]]$Driver.A)
  density.driver.A = density(driver.A, from=min(driver.A), to=max(driver.A), n=100, bw=max(driver.A)/100)
  density.driver.A.y = (density.driver.A$y - min(density.driver.A$y)) / (max(density.driver.A$y) - min(density.driver.A$y))
  driver.A.range = seq(min(driver.A), max(driver.A), length.out = 100)
  niche.A = dnorm(x=driver.A.range, mean=parameters.plot[i, "niche.A.mean"], sd=parameters.plot[i, "niche.A.sd"])
  niche.A = niche.A / max(niche.A)
  
  #computing biomass
  age = seq(0, parameters.plot[i, "maximum.age"], length.out = 100)
  biomass =  parameters.plot[i, "maximum.biomass"] / (1 +  parameters.plot[i, "maximum.biomass"] * exp(-  parameters.plot[i, "growth.rate"] * age))
  
  #filling temp dataframe
  plot.df.temp = data.frame(Species=rep(paste(parameters.plot[i, "label"], sep=""), 100),
                            Driver=rep("Driver", 100),
                            Driver.density.x=density.driver.A$x,
                            Driver.density.y=density.driver.A.y,
                            Value=driver.A.range, 
                            Suitability=niche.A, 
                            Age=age, 
                            Biomass=biomass, 
                            Reproductive.age=rep(parameters.plot[i, "reproductive.age"], 100),
                            Fecundity=rep(parameters.plot[i, "fecundity"], 100))
  
  #putting together with main dataframe
  plot.df = rbind(plot.df, plot.df.temp)
  
}#end of iterations

plot.df$Suitability = round(plot.df$Suitability, 2)
plot.df[plot.df$Suitability==0, "Suitability"]=NA
plot.df$Species<-as.character(plot.df$Species)
plot.df.niche<-plot.df[plot.df$Species %in% unique(plot.df$Species)[1:4], ]

#niche plot
plot.niche = ggplot(data=plot.df.niche, aes(x=Value, y=Suitability, group=Species)) + 
  geom_ribbon(data=plot.df.niche, aes(ymin=0, ymax=Driver.density.y), color=col.gray.light, fill=col.green, alpha=0.5, size=0.2) +
  geom_ribbon(data=plot.df.niche, aes(ymin=0, ymax=Suitability), alpha=0.8, colour=NA, fill=col.blue, size=0.2) +
  geom_line(data=plot.df.niche, aes(x=Value, y=Driver.density.y), color=col.green, alpha=0.5, size=0.2) +
  facet_wrap("Species", ncol=1) + 
  xlab("Driver values") + 
  ylab("Environmental suitability") + 
  theme(legend.position="none",
        panel.spacing = unit(0, "lines"),
        axis.text.x=element_text(size=axis.text.size - 1),
        axis.text.y=element_text(size=axis.text.size - 1),
        axis.title.x=element_text(size=axis.title.size +2),
        axis.title.y=element_text(size=axis.title.size +2),
        plot.margin = unit(c(1.7,-0.1,0,0), "cm"),
        strip.text.x = element_text(size = axis.text.size/1.1),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm")) +
  scale_x_continuous(breaks=c(0, 50, 100))


#suitability plot
suitablity.50.10 = dnorm(x=driver.A, mean=50, sd=10)
suitablity.50.10 = suitablity.50.10 / max(suitablity.50.10)
suitablity.50.20 = dnorm(x=driver.A, mean=50, sd=20)
suitablity.50.20 = suitablity.50.20 / max(suitablity.50.20)
suitablity.75.10 = dnorm(x=driver.A, mean=75, sd=10)
suitablity.75.10 = suitablity.75.10 / max(suitablity.75.10)
suitablity.75.20 = dnorm(x=driver.A, mean=75, sd=20)
suitablity.75.20 = suitablity.75.20 / max(suitablity.75.20)
suitability.df <- data.frame(time=round(seq(1, 10000, length.out = length(driver.A)), 0), Driver=driver.A, "m 50, sd 10"=suitablity.50.10, "m 50, sd 20"=suitablity.50.20, "m 75, sd 10"=suitablity.75.10, "m 75, sd 20"=suitablity.75.20)
suitability.df.long <- gather(suitability.df, variable, value, 2:6)
suitability.df.long$color <- col.blue
suitability.df.long[suitability.df.long$variable=="Driver", "color"] <- col.green

plot.suitability<-ggplot(data=suitability.df.long, aes(x=time, y=value, group=variable)) + 
  geom_line(colour = suitability.df.long$color, size=0.2) + 
  facet_wrap("variable", ncol=1, scales="free_y") +
  ylab("") + 
  xlab("Time (years)") +
  theme(legend.position="none",
        panel.spacing = unit(0, "lines"),
        axis.text.x=element_text(size=axis.text.size - 1),
        axis.text.y=element_text(size=axis.text.size - 1),
        axis.title.x=element_text(size=axis.title.size +2),
        axis.title.y=element_text(size=axis.title.size +2),
        plot.margin = unit(c(0.6,0.5,0,-0.1), "cm"),
        strip.text.x = element_text(color = "white", size = axis.text.size/1.1),
        strip.background = element_rect(fill="white"),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm")) +
        coord_cartesian(xlim=c(4500, 6500))


plot.niche.suitability <- plot_grid(plot.niche, plot.suitability, ncol=2, align="h", labels = c("a", "b"), rel_widths = c(0.4, 0.6), axis = "blr", label_size = 10)

#growth plot
plot.df.growth<-plot.df[plot.df$Species %in% unique(plot.df$Species)[5:6], ]
growth.plot.temp = ggplot(data=plot.df.growth, aes(x=Age, y=Biomass, group=Species)) + 
  geom_ribbon(ymin=0, ymax=plot.df.growth$Biomass, color="gray80", fill="gray80", alpha=0.5, size=0.2) + 
  geom_line(aes(x=Reproductive.age, y=Biomass), color=col.blue, size=1, alpha=0.8) + 
  facet_wrap(facets="Species", ncol=2, strip.position="top", scales="free_x") + 
  xlab("Age of individuals (years)") + 
  ylab("Biomass") + 
  theme(legend.position="none",
        panel.spacing = unit(1, "lines"),
        axis.text.x=element_text(size=axis.text.size - 1),
        axis.text.y=element_text(size=axis.text.size - 1),
        axis.title.x=element_text(size=axis.title.size +2),
        axis.title.y=element_text(size=axis.title.size+2),
        plot.margin = unit(c(0.2,0.5,0,0), "cm"),
        strip.text.x = element_text(size = axis.text.size/1.1),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm"))

growth.plot <- plot_grid(growth.plot.temp, labels = "c", label_size = 10)

plot_grid(plot.niche.suitability, growth.plot, rel_heights = c(figure.width, 1), ncol=1) 

ggsave(filename = "Figure_3.pdf", width=figure.width, height=figure.width + 1)

#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])

```


\newpage

## Figure 2

```{r, echo=FALSE, fig.height=1.75, fig.width=2.75, cache=FALSE, fig.cap="Example of ecological memory pattern showing the relative importance of each component at each time lag as computed by Random Forest. The ecological memory features measured are: a) strength of the concurrent effect (highlighted by a gray box); b) strength of the exogenous memory; c) strength of the endogenous component; d) length of the exogenous component; e) length of the endogenous component; f) dominance of the endogenous component; g) dominance of the exogenous component. Note that only data above the median of the Random variable (yellow line) are considered for the computations."}


load("simtable.RData")
x=E2.driver.df
x<-x[x$label==unique(x$label)[1], ]
x<-x[x$sampling==unique(x$sampling)[1], ]
x<-x[, c("median", "min", "max", "Variable", "Lag")]
x$Variable<-as.character(x$Variable)

#modifying values
x[x$Variable=="Response", "Variable"]<-"Endogenous"
x[x$Variable=="Driver.A", "Variable"]<-"Exogenous"
x[x$Variable=="Endogenous"|x$Variable=="Exogenous", "median"] <- x[x$Variable=="Endogenous"|x$Variable=="Exogenous", "median"] - 10
x[x$Variable=="Endogenous"|x$Variable=="Exogenous", "min"] <- x[x$Variable=="Endogenous"|x$Variable=="Exogenous", "min"] - 10
x[x$Variable=="Endogenous"|x$Variable=="Exogenous", "max"] <- x[x$Variable=="Endogenous"|x$Variable=="Exogenous", "max"] - 10

x[x$Variable=="Endogenous" & x$Lag >=120, "median"] <- x[x$Variable=="Endogenous" & x$Lag >=120, "median"] - c(1,2,2.5,3,4,6,7)
x[x$Variable=="Endogenous" & x$Lag >=120, "min"] <- x[x$Variable=="Endogenous" & x$Lag >=120, "min"] - c(1,2,2.5,3,4,6,7)
x[x$Variable=="Endogenous" & x$Lag >=120, "max"] <- x[x$Variable=="Endogenous" & x$Lag >=120, "max"] - c(1,2,2.5,3,4,6,7)

x[x$Variable=="Exogenous" & x$Lag >=120, "median"] <- x[x$Variable=="Exogenous" & x$Lag >=120, "median"] - c(1,1.5,2,3.5,4,4,4.5)
x[x$Variable=="Exogenous" & x$Lag >=120, "min"] <- x[x$Variable=="Exogenous" & x$Lag >=120, "min"] - c(1,1.5,2,3.5,4,4,4.5)
x[x$Variable=="Exogenous" & x$Lag >=120, "max"] <- x[x$Variable=="Exogenous" & x$Lag >=120, "max"] - c(1,1.5,2,3.5,4,4,4.5)


x[x$Variable=="Exogenous" & x$Lag <=60, "median"] <- x[x$Variable=="Exogenous" & x$Lag <=60, "median"] -5
x[x$Variable=="Exogenous" & x$Lag <=60, "min"] <- x[x$Variable=="Exogenous" & x$Lag <=60, "min"] - 5
x[x$Variable=="Exogenous" & x$Lag <=60, "max"] <- x[x$Variable=="Exogenous" & x$Lag <=60, "max"] - 5

x[x$Variable=="Exogenous" & x$Lag >=180 , "median"] <- x[x$Variable=="Exogenous" & x$Lag >=180 , "median"] - c(1,1.5,2,2)
x[x$Variable=="Exogenous" & x$Lag >=180 , "min"] <- x[x$Variable=="Exogenous" & x$Lag >=180 , "min"] - c(1,1.5,2,2)
x[x$Variable=="Exogenous" & x$Lag >=180 , "max"] <- x[x$Variable=="Exogenous" & x$Lag >=180 , "max"] - c(1,1.5,2,2)

x[x$Variable=="Exogenous" & x$Lag == 0 , "median"] <- x[x$Variable=="Exogenous" & x$Lag == 0 , "median"] - 4
x[x$Variable=="Exogenous" & x$Lag == 0 , "min"] <- x[x$Variable=="Exogenous" & x$Lag == 0 , "min"] -4
x[x$Variable=="Exogenous" & x$Lag == 0 , "max"] <- x[x$Variable=="Exogenous" & x$Lag == 0 , "max"] - 4

x[x$Variable=="Endogenous" & x$Lag <=80, "median"] <- x[x$Variable=="Endogenous" & x$Lag <=80, "median"] + c(0, 5, 4, 3)
x[x$Variable=="Endogenous" & x$Lag <=80, "min"] <- x[x$Variable=="Endogenous" & x$Lag <=80, "min"] + c(0, 5, 4, 3)
x[x$Variable=="Endogenous" & x$Lag <=80, "max"] <- x[x$Variable=="Endogenous" & x$Lag <=80, "max"] + c(0, 5, 4, 3)

x[x$Variable=="Random", "median"] <- x[x$Variable=="Random", "median"] + 8
x[x$Variable=="Random", "max"] <- x[x$Variable=="Random", "max"] + 9

annotation.size=6
legend.title.size=6
legend.text.size=6
axis.text.size=6
axis.title.size=8
annotation.factor=2.2


#plot
segment.size=0.3
ggplot(data=x, aes(x=Lag, y=median, group=Variable, color=Variable, fill=Variable)) +
  geom_ribbon(aes(ymin=min, ymax=max), alpha=0.3, colour=NA, size=0.2) + 
  geom_line(alpha=1, size=0.5) +
  scale_color_manual(values=c(col.blue, col.green, "#FEEE62")) +
  scale_fill_manual(values=c(col.blue, col.green, "#FEEE62")) +
  scale_x_continuous(breaks=unique(x$Lag)) +
  xlab("Lag (years)") + 
  ylab("Relative importance") +
  coord_cartesian(xlim=c(0, 240), ylim=c(0, 35)) +
  annotate("rect", xmin = -5, xmax = 5, ymin = 25, ymax = 31, alpha = .2) +
  annotate("text", x=30, y=33, label="Concurrent effect", color="gray40", size=annotation.size/3) + 
  annotate("segment", x = 0, xend = 0, y = 8.2, yend = 28, colour = "gray40", size=segment.size/annotation.factor) +
  annotate("text", x=-5, y=18, label="a", color="gray40", size=annotation.size/annotation.factor) +
  annotate("segment", x = 19, xend = 19, y = 8.2, yend = 18.5, colour = "gray40", size=segment.size) +
  annotate("text", x=15, y=15, label="b", color="gray40", size=annotation.size/annotation.factor) +
  annotate("segment", x = 21, xend = 21, y = 8.2, yend = 22.5, colour = "gray40", size=segment.size) +
  annotate("text", x=25, y=13, label="c", color="gray40", size=annotation.size/annotation.factor) +
  annotate("segment", x = 20, xend = 170, y = 30, yend = 30, colour = "gray40", size=segment.size) +
  annotate("text", x=90, y=32, label="d", color="gray40", size=annotation.size/annotation.factor) +
  annotate("segment", x = 20, xend = 95, y = 25, yend = 25, colour = "gray40", size=segment.size) +
  annotate("text", x=60, y=27, label="e", color="gray40", size=annotation.size/annotation.factor) +
  annotate("segment", x = 20, xend = 83, y = 3, yend = 3, colour = "gray40", size=segment.size) +
  annotate("text", x=55, y=5, label="f", color="gray40", size=annotation.size/annotation.factor) +
  annotate("segment", x = 83, xend = 170, y = 15, yend = 15, colour = "gray40", size=segment.size) +
  annotate("text", x=125, y=18, label="g", color="gray40", size=annotation.size/annotation.factor) +
  theme(legend.position="bottom",
      axis.text.x=element_text(size=axis.text.size),
      axis.text.y=element_text(size=axis.text.size),
      axis.title.x=element_text(size=axis.title.size),
      axis.title.y=element_text(size=axis.title.size),
      plot.margin = unit(c(0.2,0.2,0.6,0.2), "cm"),
      legend.margin=margin(c(-0.8, 0.3, -1, -1), unit = "lines"),
      legend.text = element_text(size=annotation.size),
      legend.title = element_text(size=annotation.size),
      legend.key.height = unit(0.5, "line"),
      legend.key.width = unit(0.5, "line"),
      axis.line.x = element_line(size=0.05),
      axis.line.y = element_line(size=0.05),
      axis.ticks.x = element_line(size=0.05),
      axis.ticks.y = element_line(size=0.05),
      axis.ticks.length=unit(.1, "cm")) +
  labs(color="Memory components:") +
  guides(fill=FALSE)
ggsave(filename = "Figure_4.pdf", width=figure.width, height=figure.width - 1)


#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])
```


## Figure (removed from paper)

```{r, echo=FALSE, fig.height=2.75, fig.width=5.5, cache=FALSE, fig.cap="Driver (light green) and a subset of simulations (marine-blue) derived from the original Annual resolution (left panel), and resampled at 2 and 10cm (centre and right panels)."}
#needs to reload data produced by Appendix I
load("Appendix_I_output.RData")

#getting time and driver for the Annual dataset
simulations<-list()
temp.time<-simulation[[1,1]]$Time
temp.time<-temp.time[temp.time >=1]
simulations[[1]]<-temp.time
temp.time<-simulation[[1,1]]$Time
temp.driver<-simulation[[1,1]]$Driver.A
temp.driver<-temp.driver[temp.time >=1]
simulations[[2]]<-temp.driver

#gathering pollen counts for every taxa
for(i in 1:16){
temp.time<-simulation[[i,1]]$Time
temp.pollen<-simulation[[i,1]]$Pollen
temp.pollen<-temp.pollen[temp.time >=1]
simulations[[i+2]]<-temp.pollen
}

#all together in a single dataframe
simulations<-data.frame(do.call("cbind", simulations))
colnames(simulations)<-c("Time", "Driver", parameters$label)

#computing correlations between driver and simulated pollen curves
correlations<-vector()
for(i in 3:18){
  correlations <- c(correlations, cor(simulations$Driver, simulations[,i]))
}

#to long format for easier plotting
simulations.long<-gather(simulations, simulation, pollen, 2:18)
simulations.long$color <- col.blue
simulations.long[simulations.long$simulation=="Driver", "color"] <- col.green

#subset
simulations.long<-simulations.long[simulations.long$simulation %in% c("Driver", "S1000A50-20_f10", "S1000A75-5_f1", "S10A50-20_f10", "S10A75-5_f1"),]
simulations.long[simulations.long$simulation=="S1000A50-20_f10", "simulation"] <- "Ls 1000, F 10, No 50, Nb 20"
simulations.long[simulations.long$simulation=="S1000A75-5_f1", "simulation"] <- "Ls 1000, F 1, No 75, Nb 10"
simulations.long[simulations.long$simulation=="S10A50-20_f10", "simulation"] <- "Ls 10, F 10, No 50, Nb 20"
simulations.long[simulations.long$simulation=="S10A75-5_f1", "simulation"] <- "Ls 10, F 1, No 75, Nb 10"

#plot 1year
sim.1cm <- ggplot(simulations.long, aes(x=Time, y=pollen, group=simulation)) +
  geom_line(colour = simulations.long$color, size=0.3) +
  facet_wrap("simulation", scales="free_y", ncol=1) +
  coord_cartesian(xlim = c(4000, 6500)) +
  theme(legend.position="none",
        axis.text.x=element_text(size=axis.text.size),
        axis.text.y=element_text(size=axis.text.size),
        axis.title.x=element_text(size=axis.title.size + 2),
        axis.title.y=element_text(size=axis.title.size + 2),
        plot.margin = unit(c(0.2,0,0,0), "cm"),
        plot.title = element_text(size=axis.title.size),
        strip.text.x = element_text(size = axis.text.size),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm"),
        panel.spacing = unit(0, "lines")) +
  ylab("Pollen concentration (blue) and driver values (green)") +
  xlab("") +
  ggtitle("Annual")

#getting simulation data for dataset 2cm
simulations<-list()
temp.time<-simulation[[1,3]]$Time
temp.time<-temp.time[temp.time >=1]
simulations[[1]]<-temp.time
temp.time<-simulation[[1,3]]$Time
temp.driver<-simulation[[1,3]]$Driver.A
temp.driver<-temp.driver[temp.time >=1]
simulations[[2]]<-temp.driver
for(i in 1:16){
  temp.time<-simulation[[i,3]]$Time
  temp.pollen<-simulation[[i,3]]$Pollen
  temp.pollen<-temp.pollen[temp.time >=1]
  simulations[[i+2]]<-temp.pollen
}

#into single dataframe
simulations<-data.frame(do.call("cbind", simulations))
colnames(simulations)<-c("Time", "Driver", parameters$label)
simulations.long<-gather(simulations, simulation, pollen, 2:18)
simulations.long$color <- col.blue
simulations.long[simulations.long$simulation=="Driver", "color"] <- col.green
#subset
simulations.long<-simulations.long[simulations.long$simulation %in% c("Driver", "S1000A50-20_f10", "S1000A75-5_f1", "S10A50-20_f10", "S10A75-5_f1"),]
simulations.long[simulations.long$simulation=="S1000A50-20_f10", "simulation"] <- "Ls 1000, F 10, No 50, Nb 20"
simulations.long[simulations.long$simulation=="S1000A75-5_f1", "simulation"] <- "Ls 1000, F 1, No 75, Nb 10"
simulations.long[simulations.long$simulation=="S10A50-20_f10", "simulation"] <- "Ls 10, F 10, No 50, Nb 20"
simulations.long[simulations.long$simulation=="S10A75-5_f1", "simulation"] <- "Ls 10, F 1, No 75, Nb 10"

#plot 2cm
sim.2cm <- ggplot(simulations.long, aes(x=Time, y=pollen, group=simulation)) +
  geom_line(colour = simulations.long$color, size=0.3) +
  facet_wrap("simulation", scales="free_y", ncol=1) +
  coord_cartesian(xlim = c(4000, 6500)) +
  theme(legend.position="none",
        axis.text.x=element_text(size=axis.text.size),
        axis.text.y=element_text(size=axis.text.size),
        axis.title.x=element_text(size=axis.title.size + 2),
        axis.title.y=element_text(size=axis.title.size + 2),
        plot.margin = unit(c(0.2,0,0,0), "cm"),
        plot.title = element_text(size=axis.title.size),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm"),
        panel.spacing = unit(0, "lines"),
        strip.text.x = element_text(color = "white", size = axis.text.size),
        strip.background = element_rect(fill="white"))  +
  ylab("") +
  xlab("Time (years)") +
  ggtitle("2cm")


#getting simulation data
simulations<-list()
temp.time<-simulation[[1,5]]$Time
temp.time<-temp.time[temp.time >=1]
simulations[[1]]<-temp.time
temp.time<-simulation[[1,5]]$Time
temp.driver<-simulation[[1,5]]$Driver.A
temp.driver<-temp.driver[temp.time >=1]
simulations[[2]]<-temp.driver
for(i in 1:16){
  temp.time<-simulation[[i,5]]$Time
  temp.pollen<-simulation[[i,5]]$Pollen
  temp.pollen<-temp.pollen[temp.time >=1]
  simulations[[i+2]]<-temp.pollen
}
simulations<-data.frame(do.call("cbind", simulations))
colnames(simulations)<-c("Time", "Driver", parameters$label)
simulations.long<-gather(simulations, simulation, pollen, 2:18)
simulations.long$color <- col.blue
simulations.long[simulations.long$simulation=="Driver", "color"] <- col.green
#subset
simulations.long<-simulations.long[simulations.long$simulation %in% c("Driver", "S1000A50-20_f10", "S1000A75-5_f1", "S10A50-20_f10", "S10A75-5_f1"),]
simulations.long[simulations.long$simulation=="S1000A50-20_f10", "simulation"] <- "Ls 1000, F 10, No 50, Nb 20"
simulations.long[simulations.long$simulation=="S1000A75-5_f1", "simulation"] <- "Ls 1000, F 1, No 75, Nb 10"
simulations.long[simulations.long$simulation=="S10A50-20_f10", "simulation"] <- "Ls 10, F 10, No 50, Nb 20"
simulations.long[simulations.long$simulation=="S10A75-5_f1", "simulation"] <- "Ls 10, F 1, No 75, Nb 10"

#plot 2cm
#plot 10cm
sim.10cm <- ggplot(simulations.long, aes(x=Time, y=pollen, group=simulation)) +
  geom_line(colour = simulations.long$color, size=0.3) +
  facet_wrap("simulation", scales="free_y", ncol=1) +
  coord_cartesian(xlim = c(4000, 6500)) +
  theme(legend.position="none",
        axis.text.x=element_text(size=axis.text.size),
        axis.text.y=element_text(size=axis.text.size),
        axis.title.x=element_text(size=axis.title.size + 2),
        axis.title.y=element_text(size=axis.title.size + 2),
        plot.margin = unit(c(0.2,0.2,0,0), "cm"),
        plot.title = element_text(size=axis.title.size),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm"),
        panel.spacing = unit(0, "lines"),
        strip.text.x = element_text(color = "white", size = axis.text.size),
        strip.background = element_rect(fill="white"))  +
  ylab("") +
  xlab("") +
  ggtitle("10cm")

plot_grid(sim.1cm, sim.2cm, sim.10cm, ncol=3, align="h") 
ggsave(filename = "Figure_5.pdf", width=figure.width*2, height=figure.width+1)

#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])

```

\newpage

## Table 1 (removed from paper)

```{r, echo=FALSE, cache=FALSE}
#preparing correlation matrix
correlation.matrix<-matrix(NA, ncol=16, nrow=5)
colnames(correlation.matrix)<-parameters$label
rownames(correlation.matrix)<-c("Annual", "1cm", "2cm", "6cm", "10cm")

#I wish I knew
for(j in 1:nrow(correlation.matrix)){

  #getting the driver for the given row
  temp.time<-simulation[[1,j]]$Time
  temp.driver<-simulation[[1,j]]$Driver.A
  driver<-temp.driver[temp.time >=1]

  simulations<-list()
  
  for(i in 1:16){
    temp.time<-simulation[[i,j]]$Time
    temp.pollen<-simulation[[i,j]]$Pollen
    temp.pollen<-temp.pollen[temp.time >=1]
    simulations[[i]]<-temp.pollen
  }
  
  simulations<-data.frame(do.call("cbind", simulations))
  colnames(simulations)<-parameters$label
  
  #correlations
  correlations<-vector()
  for(i in 1:16){
    correlation.matrix[j,i] <- cor(driver, simulations[,i])
  }
}

#preparing the correlation matrix
correlation.matrix <- data.frame(t(round(correlation.matrix, 2)))
correlation.matrix <- data.frame(maximum.age=parameters$maximum.age, fecundity=parameters$fecundity, niche.optimum=parameters$niche.A.mean, niche.breadth=parameters$niche.A.sd, correlation.matrix)
correlation.matrix <- correlation.matrix[with(correlation.matrix, order(maximum.age, fecundity, niche.optimum, niche.breadth)),]
rownames(correlation.matrix)<-NULL
correlation.matrix<-abs(correlation.matrix)
correlation.matrix <- correlation.matrix[order(c(correlation.matrix$niche.optimum, correlation.matrix$niche.breadth, correlation.matrix$maximum.age, correlation.matrix$fecundity), decreasing=FALSE), ]
correlation.matrix <- na.omit(correlation.matrix)
rownames(correlation.matrix)<-NULL

#export table
kable(correlation.matrix, format="html", row.names=NA, col.names = c("Life-span", "Fecundity", "Niche optimum", "Niche breadth", "Annual", "1cm", "2cm", "6cm", "10cm"), caption="Pearson correlation index between simulated pollen curves and the driver at different sampling resolutions (columns Annual, 1cm, 2cm, 6cm, and 10cm), compared with niche and life traits used for the simulations.", booktabs = T)  %>% 
  kable_styling(full_width = F, latex_options = c("hold_position", "striped")) %>%
  column_spec(1:9, bold = F, color="#38598C") %>%
  row_spec(0, bold=TRUE, color = "#585858") %>%
  cat(., file = "table2.html")

#print table
kable(correlation.matrix, row.names=NA, col.names = c("Life-span", "Fecundity", "Niche optimum", "Niche breadth", "Annual", "1cm", "2cm", "6cm", "10cm"), caption="Pearson correlation index between simulated pollen curves and the driver at different sampling resolutions (columns Annual, 1cm, 2cm, 6cm, and 10cm), compared with niche and life traits used for the simulations.", booktabs = T)  %>% 
  kable_styling(full_width = F, latex_options = c("hold_position", "striped")) %>%
  column_spec(1:9, bold = F, color="#38598C") %>%
  row_spec(0, bold=TRUE, color = "#585858")


#removing objects not needed any longer
# rm(list= ls()[!(ls() %in% to.preserve)])
```

\newpage

## Figure 3

```{r, echo=FALSE, fig.height=2.75, fig.width=2.75, cache=FALSE, fig.cap="Effect of variable sediment accumulation rate on ecological memory features. Dots represent virtual taxa. Values of the x axis show the difference in ecological memory features when substracting those of the 1cm dataset (variable accumulation rate) to those of the Annual dataset (constant accumulation rate). Violin-plots, in blue, show the underlying distribution of the data. Vertical blue lines represent the median."}

#getting features
E1.driver.sampling <- E1.driver.df[E1.driver.df$sampling %in% c("Annual", "1cm"), ]
E1.driver.sampling$Variable=as.character(E1.driver.sampling$Variable)
E1.driver.sampling[E1.driver.sampling$Variable=="Response", "Variable"]="Endogenous"
E1.driver.sampling[E1.driver.sampling$Variable=="Driver.A", "Variable"]="Exogenous"

#separating datasets
xAnnual <- extractMemoryFeatures(memory.pattern=E1.driver.sampling,
                              exogenous.component="Exogenous", 
                              endogenous.component="Endogenous", 
                              sampling.subset="Annual",
                              scale.strength = FALSE)

x1cm <- extractMemoryFeatures(memory.pattern=E1.driver.sampling,
                              exogenous.component="Exogenous", 
                              endogenous.component="Endogenous", 
                              sampling.subset="1cm",
                              scale.strength = FALSE)

#putting datasets together to scale strength
xAnnual$id <- "Annual"
x1cm$id <- "1cm"
temp <- rbind(xAnnual, x1cm)

#scaling strength between 0 and 1
temp$strength.concurrent <- temp$strength.concurrent / max(temp$strength.concurrent)
temp$strength.exogenous <- temp$strength.exogenous / max(temp$strength.exogenous)
temp$strength.endogenous <- temp$strength.endogenous / max(temp$strength.endogenous)

#separatind datasets
xAnnual <- temp[temp$id == "Annual", ]
x1cm <- temp[temp$id == "1cm", ]


#differences between datasets
xDiff <- xAnnual
for(i in 2:8){
  xDiff[,i] <- x1cm[,i] - xAnnual[,i]
}

#to long format
xDiff.long <- gather(xDiff, mem.feature.name, mem.feature.value, 2:8)

#tidy names
xDiff.long[xDiff.long$mem.feature.name=="dominance.endogenous", "mem.feature.name"]<-"Dominance - endogenous"
xDiff.long[xDiff.long$mem.feature.name=="dominance.exogenous", "mem.feature.name"]<-"Dominance - exogenous"
xDiff.long[xDiff.long$mem.feature.name=="length.endogenous", "mem.feature.name"]<-"Length - endogenous"
xDiff.long[xDiff.long$mem.feature.name=="length.exogenous", "mem.feature.name"]<-"Length - exogenous"
xDiff.long[xDiff.long$mem.feature.name=="strength.concurrent", "mem.feature.name"]<-"Strength - concurrent"
xDiff.long[xDiff.long$mem.feature.name=="strength.endogenous", "mem.feature.name"]<-"Strength - endogenous"
xDiff.long[xDiff.long$mem.feature.name=="strength.exogenous", "mem.feature.name"]<-"Strength - exogenous"

#factor order
xDiff.long$mem.feature.name <- factor(xDiff.long$mem.feature.name, levels=rev(c("Strength - concurrent", "Strength - endogenous", "Length - endogenous", "Dominance - endogenous", "Strength - exogenous", "Length - exogenous", "Dominance - exogenous")))

#plot
ggplot(xDiff.long, aes(y=mem.feature.value, x=as.factor(mem.feature.name))) + 
  geom_hline(yintercept = 0, color = "gray80", size=0.5, alpha=0.5) +
  geom_violin(draw_quantiles = 0.5, color="gray50", fill = "transparent", size=0.2) +
  geom_point(size=1.5, alpha=0.7, position = position_jitter(w = 0.3, h = 0), fill = col.green, shape=21, color="gray20", stroke = 0.1) +
  xlab("") +
  ylab("") +
  theme(plot.margin = unit(c(0.2, 0.2, 0.2, -0.6), "cm"),
        legend.position="none",
        axis.text.x=element_text(size=axis.text.size),
        axis.text.y=element_text(size=axis.text.size +2),
        axis.title.x=element_text(size=axis.title.size + 2),
        axis.title.y=element_text(size=axis.title.size + 2),
        plot.title = element_text(size=axis.title.size),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm")) +
  ylab("Difference induced by \n accumulation rate") +
  coord_flip()

ggsave(filename = "Figure_6.pdf", width=figure.width, height=figure.width)

#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])
```

\newpage

## Figure 4

```{r, echo=FALSE, fig.height=4.75, message=FALSE, warning=FALSE, error=FALSE, fig.width=2.75, cache=FALSE, fig.cap="Ecological memory features across decreasing sampling densities. Dots represent virtual taxa (horizontal jitter was applied to the points for ease of examination)."}

#getting features
E1.driver.sampling <- E1.driver.df[E1.driver.df$sampling %in% c("1cm", "2cm", "6cm", "10cm"), ]
E1.driver.sampling$Variable=as.character(E1.driver.sampling$Variable)
E1.driver.sampling[E1.driver.sampling$Variable=="Response", "Variable"]="Endogenous"
E1.driver.sampling[E1.driver.sampling$Variable=="Driver.A", "Variable"]="Exogenous"
E1.driver.sampling$sampling<-factor(E1.driver.sampling$sampling, levels = c("1cm", "2cm", "6cm", "10cm"))

#separating datasets
x1cm <- extractMemoryFeatures(memory.pattern=E1.driver.sampling,
                              exogenous.component="Exogenous", 
                              endogenous.component="Endogenous", 
                              sampling.subset="1cm",
                              scale.strength = FALSE)
x1cm$cm=1

x2cm <- extractMemoryFeatures(memory.pattern=E1.driver.sampling,
                              exogenous.component="Exogenous",
                              endogenous.component="Endogenous",
                              sampling.subset="2cm",
                              scale.strength = FALSE)
x2cm$cm=2

x6cm <- extractMemoryFeatures(memory.pattern=E1.driver.sampling,
                              exogenous.component="Exogenous",
                              endogenous.component="Endogenous",
                              sampling.subset="6cm",
                              scale.strength = FALSE)
x6cm$cm=6

x10cm <- extractMemoryFeatures(memory.pattern=E1.driver.sampling,
                               exogenous.component="Exogenous",
                               endogenous.component="Endogenous",
                               sampling.subset="10cm",
                              scale.strength = FALSE)
x10cm$cm=10

#into a single dataframe
x.cm <- rbind(x1cm, x2cm, x6cm, x10cm)

#scaling strength
#scaling strength between 0 and 1
x.cm$strength.concurrent <- x.cm$strength.concurrent / max(x.cm$strength.concurrent)
x.cm$strength.exogenous <- x.cm$strength.exogenous / max(x.cm$strength.exogenous)
x.cm$strength.endogenous <- x.cm$strength.endogenous / max(x.cm$strength.endogenous)

#long format for plotting
x.cm.long <- gather(x.cm, feature, value, 2:8)
x.cm.long$niche.mean <- factor(x.cm.long$niche.mean)
x.cm.long$niche.sd <- factor(x.cm.long$niche.sd)
x.cm.long$maximum.age <- factor(x.cm.long$maximum.age)
x.cm.long$fecundity <- factor(x.cm.long$fecundity)

#yada yada yada
x.cm.long$feature<-as.character(x.cm.long$feature)
x.cm.long[x.cm.long$feature=="dominance.endogenous", "feature"]<-"Dominance - endogenous"
x.cm.long[x.cm.long$feature=="dominance.exogenous", "feature"]<-"Dominance - exogenous"
x.cm.long[x.cm.long$feature=="length.endogenous", "feature"]<-"Length - endogenous"
x.cm.long[x.cm.long$feature=="length.exogenous", "feature"]<-"Length - exogenous"
x.cm.long[x.cm.long$feature=="strength.concurrent", "feature"]<-"Strength - concurrent"
x.cm.long[x.cm.long$feature=="strength.endogenous", "feature"]<-"Strength - endogenous"
x.cm.long[x.cm.long$feature=="strength.exogenous", "feature"]<-"Strength - exogenous"

#factor order
x.cm.long$feature <- factor(x.cm.long$feature, levels=c("Strength - concurrent", "Strength - endogenous", "Length - endogenous", "Dominance - endogenous", "Strength - exogenous", "Length - exogenous", "Dominance - exogenous"))

#yeah, the plot
ggplot(x.cm.long, aes(x=cm, y=value)) +
  geom_point(size=1, position = position_jitter(w = 0.2, h = 0.01), fill = col.green, shape=21, color="gray20", stroke = 0.1) +
  # geom_point(aes(group=cm), shape=21, color="gray20", alpha=0.5, position=position_jitter(width = 0.2, height = 0)) +
  facet_wrap("feature", ncol=1) +
  scale_x_continuous(breaks=c(1,2,6,10)) +
  scale_fill_manual(values=c(col.green, "#FEEE62")) +
  geom_smooth(aes(group=feature), method = lm, se = TRUE, color=col.blue, size=0.5) +
  xlab("Pollen sampling interval (cm)") +
  ylab("") +
  labs(fill="Niche optimum", size="Maximum age") +
  theme(axis.text.x = element_text(size=12), legend.text=element_text(size=8), legend.title = element_text(size=10)) +
  guides(fill = guide_legend(override.aes = list(size = 5))) + 
  theme(legend.position="bottom",
      axis.text.x=element_text(size=axis.text.size +2),
      axis.text.y=element_text(size=axis.text.size),
      axis.title.x=element_text(size=axis.title.size + 2),
      axis.title.y=element_text(size=axis.title.size),
      plot.margin = unit(c(0.2,0.2,0,-0.2), "cm"),
      strip.text.x = element_text(size = axis.title.size, margin=margin(0,0,0.05,0, "cm")),
      panel.spacing = unit(0, "lines"),
      axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm"),
      panel.grid.major.y = element_line( size=.1, color="gray80" ))

ggsave(filename = "Figure_7.pdf", width=figure.width, height=figure.width + 2)

#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])
```

\newpage

## Figure 5

```{r, echo=FALSE, fig.height=3.5, message=FALSE, warning=FALSE, error=FALSE, fig.width=3.625, cache=FALSE, fig.cap="Change in values of the different ecological memory features when using driver as exogenous memory component instead of suitability. Colors represent niche breadth (higher if blue), and size represent niche position (larger for a marginal niche position)."}

#extracting memory features of the suitability data
features.suitability = extractMemoryFeatures(memory.pattern = E1.suitability.df, 
                                             exogenous.component = "Suitability", 
                                             endogenous.component = "Response", 
                                             sampling.subset = "Annual", 
                                             scale.strength = FALSE)
features.suitability$experiment="suitability"

#extracting memory features of the driver data
features.driver = extractMemoryFeatures(memory.pattern = E1.driver.df,
                                        exogenous.component = "Driver.A", 
                                        endogenous.component = "Response", 
                                        sampling.subset = "Annual", 
                                        scale.strength = FALSE)
features.driver$experiment="driver"

#together to scale strength
temp <- rbind(features.suitability, features.driver)

#scaling strength between 0 and 1
temp$strength.concurrent <- temp$strength.concurrent / max(temp$strength.concurrent)
temp$strength.exogenous <- temp$strength.exogenous / max(temp$strength.exogenous)
temp$strength.endogenous <- temp$strength.endogenous / max(temp$strength.endogenous)

#separatind datasets
features.suitability <- temp[temp$experiment == "suitability", ]
features.driver <- temp[temp$experiment == "driver", ]


#differences between datasets: DRIVER - SUITABILITY
differences <- features.driver
for(i in 2:8){
  differences[,i]<- features.driver[,i] - features.suitability[,i]
}
differences$niche.mean=as.factor(differences$niche.mean)
differences$niche.sd=as.factor(differences$niche.sd)
differences$maximum.age=as.factor(differences$maximum.age)

#identifying 

#to long
differences.long <- gather(differences, mem.feature.name, mem.feature.value, 2:8)

#factors
differences.long$maximum.age <- as.factor(differences.long$maximum.age )
differences.long$fecundity <- as.factor(differences.long$fecundity )
differences.long$niche.mean <- as.factor(differences.long$niche.mean )
differences.long$niche.sd <- as.factor(differences.long$niche.sd )

#tidy names
differences.long[differences.long$mem.feature.name=="dominance.endogenous", "mem.feature.name"]<-"Dominance - endogenous"
differences.long[differences.long$mem.feature.name=="dominance.exogenous", "mem.feature.name"]<-"Dominance - exogenous"
differences.long[differences.long$mem.feature.name=="length.endogenous", "mem.feature.name"]<-"Length - endogenous"
differences.long[differences.long$mem.feature.name=="length.exogenous", "mem.feature.name"]<-"Length - exogenous"
differences.long[differences.long$mem.feature.name=="strength.concurrent", "mem.feature.name"]<-"Strength - concurrent"
differences.long[differences.long$mem.feature.name=="strength.endogenous", "mem.feature.name"]<-"Strength - endogenous"
differences.long[differences.long$mem.feature.name=="strength.exogenous", "mem.feature.name"]<-"Strength - exogenous"

#factor order
differences.long$mem.feature.name <- factor(differences.long$mem.feature.name, levels=rev(c("Strength - concurrent", "Strength - endogenous", "Length - endogenous", "Dominance - endogenous", "Strength - exogenous", "Length - exogenous", "Dominance - exogenous")))



#plot
ggplot(differences.long, aes(y=mem.feature.value, x=mem.feature.name)) + 
  geom_hline(yintercept = 0, color = "gray80", size=0.5, alpha=0.5) +
  geom_violin(draw_quantiles = 0.5, color="gray50", fill = "transparent", size=0.2) +
  geom_point(aes(fill=niche.sd, size=niche.mean), position = position_jitter(w = 0.3, h = 0), shape=21, color="gray20", stroke = 0.1, alpha=0.6) +
    scale_fill_manual(values=c(col.green, col.blue)) +
    scale_size_discrete(range = c(1, 2)) +
  xlab("") +
  ylab("") +
  theme(plot.margin = unit(c(0.2, 0.3, 0.2, -0.5), "cm"),
        legend.position="bottom",
        axis.text.x=element_text(size=axis.text.size +2),
        axis.text.y=element_text(size=axis.text.size),
        axis.title.x=element_text(size=axis.title.size + 2),
        axis.title.y=element_text(size=axis.title.size + 2),
        plot.title = element_text(size=axis.title.size),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm"),
        legend.text = element_text(size = axis.text.size),
        legend.title = element_text(size = axis.text.size),
        legend.key.width = unit(0.01, "line"),
        legend.margin=margin(c(-0.4, 0, 0, 0), unit = "lines"),
        legend.justification = "left", 
        legend.spacing.y = unit(0.1, "cm"),
        legend.box = "vertical") +
    ylab("Difference when using driver \n  instead of suitability") + 
    guides(fill = guide_legend(override.aes = list(size = 2))) +
    labs(fill = "Niche breadth", size="Niche optimum") +
  coord_flip()

ggsave("Figure_8.pdf", width=figure.width, height=figure.width + 1)

#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])
```


## Figure 6

```{r, echo=FALSE, fig.height=2.25, message=FALSE, warning=FALSE, error=FALSE, fig.width=2.75, cache=FALSE, fig.cap="Relationship between life traits, niche features, and ecological memory components, when using environmental suitability as exogenous component. Colors represent niche breadth, size represents niche optimum, and shape represents life-span. Numbers within figures represent fecundity. X and Y axes are computed as the average of the length, dominance, and strength of the endogenous and exogenous components of ecological memory. Contour lines represent the relative strength of the concurrent effect (synchronous influence of the driver over the suitability response)."}

#getting Annual dataset
E1.driver.1cm <- E1.suitability.df[E1.suitability.df$sampling=="Annual", ]
E1.driver.1cm$Variable <- as.character(E1.driver.1cm$Variable)
E1.driver.1cm[E1.driver.1cm$Variable=="Response", "Variable"]="Endogenous"
E1.driver.1cm[E1.driver.1cm$Variable=="Suitability", "Variable"]="Exogenous"

#extracting memory features
mem.features <- extractMemoryFeatures(memory.pattern=E1.driver.1cm, 
                                      exogenous.component="Exogenous", 
                                      endogenous.component="Endogenous", 
                                      sampling.subset="Annual", 
                                      scale.strength = TRUE)
mem.features$sampling <- NULL

#mean of endogenous and exogenous components
mem.features$endogenous.memory <- rowMeans(mem.features[, c("strength.endogenous", "length.endogenous", "dominance.endogenous")])
mem.features$exogenous.memory <- rowMeans(mem.features[, c("strength.exogenous", "length.exogenous", "dominance.exogenous")])

mem.features$difference <- mem.features$endogenous.memory / mem.features$exogenous.memory


#traits to factor
mem.features$maximum.age <- as.factor(mem.features$maximum.age)
mem.features$fecundity <- as.factor(mem.features$fecundity)
mem.features$niche.mean <- as.factor(mem.features$niche.mean)
mem.features$niche.sd <- as.factor(mem.features$niche.sd)

#interpolating concurrent.effect to be plotted as contour
glm.model <- glm(strength.concurrent ~ poly(endogenous.memory, 2, raw=TRUE) + poly(exogenous.memory, 3, raw=TRUE), data = mem.features)

#pseudo R-squared
# cor(predict(glm.model), mem.features$strength.concurrent)
# summary(glm.model)
# cor(mem.features[,c("endogenous.memory", "exogenous.memory")])

#generating grid of data to predict the interpolation
xgrid <-  seq(min(mem.features$endogenous.memory), max(mem.features$endogenous.memory), 0.01)
ygrid <-  seq(min(mem.features$exogenous.memory), max(mem.features$exogenous.memory), 0.01)
grid <-  expand.grid(endogenous.memory = xgrid, exogenous.memory = ygrid)
prediction <- predict(glm.model, newdata = grid)
prediction.df <- data.frame(endogenous.memory=grid$endogenous.memory, exogenous.memory=grid$exogenous.memory, concurrent.effect=prediction)

#limiting extrapolation
prediction.df <- prediction.df[prediction.df$concurrent.effect >= 0.45, ]

#scatterplot
ggplot(data=mem.features, aes(x=endogenous.memory, y=exogenous.memory)) + 
  geom_contour(data=prediction.df, aes(z=concurrent.effect, alpha = stat(level), label = stat(level)), color="gray20", size=0.3, breaks=c(0.6, 0.7, 0.8, 0.9, 1, 1.2)) +
  scale_alpha(range=c(0.2, 1), breaks=c(0.6, 0.8, 1)) +
  scale_size_discrete(range = c(3, 4.5)) +
  scale_shape_manual(values=c(21,24)) +
  geom_point(aes(size=niche.mean, shape=maximum.age, fill=niche.sd), stroke=0.1, alpha=0.7) +
  geom_text(aes(label=fecundity), hjust=0.5, vjust=0.5, size=1.3, color="black") +
  scale_fill_manual(values=c(col.green, col.blue)) +
  scale_color_manual(values=c(col.green, col.blue)) +
  guides(text=FALSE, 
         alpha=guide_legend(order = 1),
         fill=guide_legend(order = 3, override.aes=list(shape=22, stroke=0)),
         size=guide_legend(order = 2, override.aes=list(shape=22)),
         color=FALSE
         ) +
  xlab("Endogenous memory") +
  ylab("Exogenous memory") +
  theme(plot.margin = unit(c(0.6, 0.2, 0, 0), "cm"),
        legend.position="right",
        axis.text.x=element_text(size=axis.text.size),
        axis.text.y=element_text(size=axis.text.size),
        axis.title.x=element_text(size=axis.title.size + 1),
        axis.title.y=element_text(size=axis.title.size + 1),
        plot.title = element_text(size=axis.title.size),
        axis.line.x = element_line(size=0.05),
        axis.line.y = element_line(size=0.05),
        axis.ticks.x = element_line(size=0.05),
        axis.ticks.y = element_line(size=0.05),
        axis.ticks.length=unit(.1, "cm"),
        legend.text = element_text(size = axis.text.size),
        legend.title = element_text(size = axis.text.size),
        legend.key.height = unit(0.01, "line"),
        legend.margin=margin(c(-0.8, 0, 0, -0.5), unit = "lines"),
        legend.justification = "left") +
  labs(alpha="Concurrent effect", fill = "Niche breadth", size="Niche optimum", shape="Life-span") +
  scale_x_continuous(breaks=seq(0, 1, by=0.1), limits=c(0.39,0.8)) +
  scale_y_continuous(breaks=seq(0, 1, by=0.1), limits = c(0.399, 1))
    
    ggsave("Figure_9.pdf", width=figure.width, height=figure.width - 0.5)


#removing objects not needed any longer
rm(list= ls()[!(ls() %in% to.preserve)])
```


```{r, echo=FALSE}

```

